---
title: "Juvenile Snowshoe Hares: Mini Exploratory Report"
author: "Katherine Rosencrance"
date: "11/7/2021"
output: 
   html_document: 
     theme: journal
---

# I. Introduction

This report explores several characteristics of juvenile snowshoe hares (*Lepus americanus*) live-trapped as part of a long term ecological research study at the Bonanza Creek Experimental Forest. Hares were trapped from 1999-2012 at 3 different sampling sites, or "grids". The sites included a flood zone along the Tanana River (Bonanza Riparian), mature stands of white spruce, paper birch, alder, black spruce, Labrador tea, willows, and prickly rose (Bonanza Mature), and a mature lowland black spruce stand (Black Spruce). Researchers recorded the date each hare was trapped as well as physical measurements for each (Kielland et al., 2017). The following sections include visualizations and analyses on juvenile hares only, comparing trap numbers per year , their weight by sex and site, and displaying the relationship between hind foot length and individual weight.

# II. Data and Analysis

Dr. Knut Kielland and colleagues collected these data to explore population densities in boreal regions of northern Alaska with the Bonanza Creek LTER (long term ecological research)(Kielland et al., 2017). The data contains physical measurements for hare ear size, weight, sex, age, and hind foot length. First, the report visualizes hares trapped per year and the weight of male hares versus female hares at each site. It goes on to compare male and female hare weight using an unpaired t-test (significance level of 0.05) and Cohen's d effect size. The relationship of hind foot length and weight is explained through a simple linear regression and Pearson's r correlation. All analyses are in R version 4.1.1 using RStudio.

# III. Juvenile Hare Initial Investigation

## A. Annual juvenile hare trap counts

```{r setup, include = FALSE, message = FALSE}

# Install packages

knitr::opts_chunk$set(echo = FALSE,message = FALSE,warning = FALSE)
library(tidyverse)
library(here)
library(lubridate)
library(plotly)
library(ggbeeswarm)
library(kableExtra)
library(effsize)
library(broom)

```

```{r}

# read in the data

bonanza_hares <- read_csv(here("data", "bonanza_hares.csv"))

```

```{r}

# Wrangle to produce graph of total number of juvenile trappings during each year of the study

## filter by age, convert dates into usable format, and add a new column with year
juvenile_hare_traps <- bonanza_hares %>% 
  filter(age == "j") %>% 
  mutate(date = mdy(date)) %>% 
  mutate(year = year(date))
  
```

```{r}
## create summary table to include count of traps by year

juvenile_table <- juvenile_hare_traps %>% 
  group_by(year) %>% 
  summarise(count = n())

```

```{r include = FALSE}

## produce graph
annual_juvenile_graph <- ggplot(data = juvenile_table, aes(x = year, y = count)) +
  geom_bar(color = "goldenrod2",
           fill = "goldenrod3",
           stat = "identity")+
  scale_x_continuous(breaks = 1999:2012)+
  scale_y_continuous(limits=c(0, 130),
                     breaks = c(0, 25, 50, 75, 100, 125))+
  theme_bw()+
  labs( x = "Year",
        y = "Number of Juvenile Hares Trapped",
        title = "Annual Juvenile Hare Trap Counts",
        subtitle = "1999-2012")+
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, face = "bold"),
        axis.title.x = element_text(vjust = 0.2, face = "bold"),
        axis.title.y = element_text(face = "bold"))+
  theme(plot.background = element_rect(fill = "floralwhite"))

annual_juvenile_graph


```

```{r}
## make it interactive
ggplotly(annual_juvenile_graph)

```

**Figure 1: Annual juvenile hare trap counts from 1999-2012.** Height of the bar indicates the amount of hares trapped that year at all 3 sites combined.

```{r, include = FALSE}
## Use table to get summary statistics for yearly traps
juvenile_summary_table <- juvenile_table %>% 
  summarise(mean_traps = mean(count),
            median_traps = median(count),
            max_traps = max(count),
            min_traps = min(count))

juvenile_summary_table

```

### Analysis

The data visualization clearly shows that there is a large variation between years in the number of juvenile hares that were collected. The max was `r juvenile_summary_table$max_traps` while the minimum was only `r juvenile_summary_table$min_traps`, excluding the years there were no data. In general, there is a larger number of hares collected in the first few years of the study than in the last. In fact, the first year of the study had the highest amount of hares trapped. The mean number of juvenile hares trapped over the 13 year period is `r juvenile_summary_table$mean_traps` while the median is only `r juvenile_summary_table$median_traps`. In moving forward, I would make sure that the people collecting data know that it is crucial to have uniform sampling effort from year to year. This means sampling at similar times each month with the same amount of traps set and the same amount of days. If the effort varies from year to year, the data will no longer reflect actual sample size.

## B. Visualize juvenile hare weights

```{r}
# Create new subset with names in observations changed to full names * note this is called "for beeswarm" but used for other analyses
juvenile_for_beeswarm <- juvenile_hare_traps %>% 
  mutate(grid = case_when(grid == "bonrip" ~ "Riparian",
    grid == "bonmat" ~ "Mature",
    grid == "bonbs" ~ "Black Spruce")) %>% 
    mutate(sex = case_when(
      sex == "f" ~ "female",
      sex == "m" ~ "male",
      is.na(sex) == TRUE ~ "unknown"))

# Visualize data where only juvenile hares are compared by sex and site
juv_sex_graph <- ggplot(data = juvenile_for_beeswarm, aes(x = sex, y = weight, color = sex)) +
  geom_beeswarm(cex = 3) +
  stat_summary(fun=mean, 
               geom="point", 
               shape=20, 
               size=4, 
               color="black", 
               fill="black")+
  scale_color_manual(values = c("goldenrod1","goldenrod4","azure4")) +
  theme_bw() +
  labs(x = "Sex",
       y = "Weight (g)",
       title = "Juvenile Hare Weight (g) by Sex and Grid") +
 theme(plot.title=element_text(hjust = 0.5, face = "bold"),
       axis.title.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
       panel.border = element_rect(colour = "black",
                                   fill=NA,
                                   size=0.5))+
  theme(plot.background = element_rect(fill = "floralwhite")) +
  facet_wrap(~grid)

 ggplotly(juv_sex_graph) 
  

```

**Figure 2: Weight observations (in grams) for juvenile hares taken at 3 different sites.** The sites include a mature lowland black spruce stand (Black Spruce), stands of mature white spruce, paper birch, and alder, as well as black spruce (Mature), and flood zone along the Tanana River (Riparian). Females are represented by the yellow points and male represented by brown. Hares trapped, but with no recorded sex, are represented as gray points. The black points represent the mean number of hares for each sex.

### Analysis

This graph displays the distribution of hare weight according to sex and site. It can be observed that there were a lot more hares trapped in the riparian site than the other two sites. In addition, one can see that at every site, the male mean is greater than the female mean, leading us to the conclusion that male hares, in accordance with normal sexual dimorphism patterns, are usually larger. It may also be important to observe that the means for all 3 sites fall within a close range of 800-1100 grams.

## C. Juvenile weight comparison (male and female snowshoe hare)

```{r}
# Create table of summary statistics for male and female hare weight

juv_summary_table <- juvenile_for_beeswarm %>%
  filter(sex != "unknown") %>% 
  group_by(sex) %>% 
  summarize(mean_weight = round(mean(weight, na.rm = TRUE),2),
    sd_weight = round(sd(weight, na.rm = TRUE),2),
    n = n()
  )

juv_summary_table %>% 
  kable(col.names = c("Sex", "Mean weight (g)", "Standard deviation (g)", "Sample size"), caption = "Table 1: Summary statistics for male and female hares captured from 1999-2012") %>% 
  kable_classic(position = "center")

```

```{r, include = FALSE}
# pull vectors of weight for male and female
m_weight <- juvenile_for_beeswarm %>% 
  filter(sex == "male") %>%
  filter(weight != "unknown") %>% 
  pull(weight)

f_weight <- juvenile_for_beeswarm %>% 
  filter(sex == "female") %>% 
  filter(weight != "unknown") %>% 
  pull(weight)

# Plot histograms and qqplots to check assumptions
hist(m_weight) # Looks relatively normal, maybe negatively skewed
qqnorm(m_weight) # Relatively linear
hist(f_weight) # Possibly a bit negatively skewed
qqnorm(f_weight) # Looks very linear

# Assumptions look correct, everything looks relatively normally distributed and linear

# Run the t test
juvenile_ttest <- t.test(f_weight, m_weight)

# Get the tidy model results to call outputs in-line
juvenile_ttest_tidy <- tidy(juvenile_ttest)

# Get the means & SDs for each: 
male_mean <- mean(m_weight)
female_mean <- mean(f_weight)
male_sd <- sd(m_weight)
female_sd <- sd(f_weight)

mean_difference <- male_mean - female_mean


# Effect size
mf_effsize <- cohen.d(m_weight, f_weight)

mf_effsize
```

### Analysis

The means of males versus females were compared using a few different statistical tests. Initially, the actual difference in mean was calculated to be `r round(mean_difference,2)` (grams). This number alone may not be enough to really tell if the means are statistically different. Upon running a t-test, the p-value returned as `r round(juvenile_ttest_tidy$p.value,3)`. This means that there is only a `r (round(juvenile_ttest_tidy$p.value,3))*100`% chance that if you take 2 samples from populations with the same means the difference would be at least this different. This would normally lead one to believe that the male and female weights are sigificantly different, however the effect size was also calculated using the Cohen's d test. This rendered an effect size of `r round(mf_effsize$estimate,2)` which is interpreted as small. While the statistical t-test renders the means different, the effect size is small meaning the difference is negligible.

## D. Relationship between juvenile weight and hind foot length

```{r}
# Do linear regression
juvenile_lm <- lm(weight ~ hindft, data = juvenile_for_beeswarm)

# Get tidy versions of the model output to call later on in text
juvenile_lm_tidy <- tidy(juvenile_lm)
juvenile_lm_glance <- glance(juvenile_lm)

# See model diagnostics:
# plot(juvenile_lm) in console

# Get Pearson's r correlation
juvenile_cor <- cor.test(juvenile_for_beeswarm$weight, juvenile_for_beeswarm$hindft)

# Tidy version of correlation output:
juvenile_cor_tidy <- tidy(juvenile_cor)
```

```{r}
# Create graph exploring correlation between hind foot length and weight in juvenile hares

hindft_weight_graph <- ggplot(data = juvenile_for_beeswarm, aes(x = hindft, y = weight))+
  geom_point(size = 2, color = "goldenrod")+
  theme_bw()+
  theme(plot.title=element_text(hjust = 0.5, face = "bold"))+
  theme(plot.subtitle=element_text(hjust = 0.5, face = "bold"))+
  theme(plot.background = element_rect(fill = "floralwhite"))+
  theme(axis.title.x = element_text(face = "bold"))+
  theme(axis.title.y = element_text(face = "bold"))+
  labs(x = "Hind foot Length (mm)", y = "Weight (g)", title = "Hind foot Length (mm) Vs Weight (g) in Juvenile Snowshoe Hares",
       subtitle = "1999-2012")+
   geom_smooth(method = "lm", se = FALSE, color = "black")

ggplotly(hindft_weight_graph)
```

**Figure 3: Hind foot length (mm) versus weight (g) in juvenile snowshoe hares.** Hind foot length is measured in mm and weight is measured in grams. Each point represents a hare observed at the specific weight and hind foot length. The black line is the line of best fit in the model.β1 = `r round(juvenile_lm_tidy$estimate[2], 2)` g mm-1, p < 0.001, R2 = `r round(juvenile_lm_glance$r.squared, 2)`, Pearson’s r = `r round(juvenile_cor_tidy$estimate,2)`)

### Analysis

The relationship between hare hind foot length and weight, upon observation, is positively correlated. After running a linear regression, one can see that the p-value returned was `r juvenile_lm_glance$p.value`. This is a very low p-value so one might conclude that the likelihood of getting two samples this different if samples were taken from populations with the same mean is very small. In addition, the estimate rendered when comparing the hind foot and weight via a Pearson's correlation was high at `r round(juvenile_cor_tidy$estimate,2)`. The r squared value of `r round(juvenile_lm_glance$r.squared, 2)` however, represents the fact that only `r (round(juvenile_lm_glance$r.squared,2))*100`% of the variation in weight is attributed to hind foot length. This means that other factors like sex and age should be taken into account when seeing what affects weight.

# IV. Summary

Exploratory data analysis findings include:

-   More hares were collected in the beginning of the study

-   Male hares tend to be bigger than female hares

-   The most hares were caught in the riparian site

-   Male and female weights are

-   The relationship between hind foot length and weight may seem positively correlated at first, but furhter analysis concludes that weight only attributes a small amount to the change in hind foot size

-   More factors need to be included in studying the relationship of hind foot size and weight

# V. Citation

**Data citation:** Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. <https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14>
